---
title: "DEMETER analysis"
author: "John Bruce"
date: "12/5/2017"
output: html_document
params:
  learning_rate: 0.005
  g_s: 4e-5
  alpha_beta_gamma: 0.9
  iterations: 10
  data_source: "data"
  source_description: "Novartis project DRIVE pooled shRNA screen"
  qc_genes: "BRAF,KRAS,PRMT5"
  remote_file_file: "s3://ps-informatics/bioinformatics/data/drive_input_gct2.rds"
  date: !r format(Sys.Date())
---

## Background

* Analysis of Drive data using Demeter https://data.mendeley.com/datasets/y3ds55n88r/4
* These are the raw data from the DRIVE pooled shRNA screen. It contains read counts for 398 cancer cell lines screened for 7837 genes with an average of 20 shRNAs per gene
* The data was preprocessed to generate drive_input_gct.rds. The original DRIVE data was converted into wide table format, with sequence+Gene symbols for rows, cell lines for columns, and log2(FC) for values. Details of this preprocessing step are given in preprocessing.Rmd notebook in the RSA folder.

## Data processing

Used rmarkdown parameters for passing variables into the document
```{r setup, echo=FALSE, results='hide'}
source("setup.R")
proj_dir <- opts_knit$get("proj.dir")
opts_knit$set(data.dir = file.path(proj_dir, params$data_source))
data_dir <- opts_knit$get("data.dir")
src_dir <- opts_knit$get("src.dir")
demeter_dir <- opts_knit$get("demeter.dir")
remote_file <- params$remote_file
#analysis <- "demeter"
local_file <- file.path(data_dir,
                        basename(remote_file))
# create data directory if it doesn't exist
#output_dir <- file.path(opts_knit$get("data.dir"),
#                        analysis,
#                        params$date)
#ifelse(!dir.exists(output_dir), 
#       dir.create(output_dir,recursive=TRUE), FALSE)
#source(file.path(proj_dir,"src","R","setup.R"))
Sys.setenv(Remote_File = remote_file,
           Data_Dir = data_dir,
           Src_Dir = src_dir,
           Demeter_Dir = demeter_dir,
           Local_File = local_file,
           #Analysis = analysis,
           Proj_Dir = proj_dir,
           UUID = uuid, 
           Run_Date = params$date)
```
+ Input data download from AWS
```{bash download_input_from_s3, echo=FALSE}
echo $Remote_File
echo $Local_File
if [ ! -f "$Local_File" ]
then
  aws s3 cp $Remote_File $Local_File
  echo 'Saved file here:' $Data_Dir
else
  echo 'File exists...'
fi

```

### Read in GCT file
+ Basic QC to check missing data
```{r read_data}
con <- gzfile(local_file,'rb')
table <- readRDS(con)
table[,1] <- paste0(table[,1],'_',table[,2])
# Rename first two column names
colnames(table)[c(1:2)] <- c("NAME","Description")
table[1:5,1:5]
# Third column not needed
table <- table[,-3]
missing_data_cols <- data.frame("sample_name"=names(table)[!!colSums(is.na(table))], 
                                stringsAsFactors = FALSE)
drop_cols <- missing_data_cols
if (length(missing_data_cols$sample_name)>0) {
  table[, drop_cols$sample_name]
  # needed to resave file as rds format
 new_name <- paste("cleaned",basename(local_file),sep="_")
 local_file <- file.path(opts_knit$get("data.dir"), 
                  new_name)
 tryCatch(fwrite(table, file=local_file, sep=" ", quote=TRUE),
        error = function(e) return(sprintf("Caught error: %s", e)))
}

#output table as gct format
header1 <- "#1.2"
nrows <- dim(table)[1]
ncols <- dim(table)[2] -2
header2 <- paste(nrows, ncols, collapse = " ")
outfile <- file.path(data_dir,"all.gct")
cat(header1, header2, file=outfile, sep="\n")
write.table(table,file=outfile, append=T, row.names=F, quote=F)
```

### Batch.csv creation
* Take the third line (head -3 | tail -1).
* Replace tab with line feed.
* Skip first two lines (tail -n +3).
* Place each word in quotes: 
```{bash batch.csv_creation}
cat << EOF > $Data_Dir/batches.csv
"Name","DEMETER batch"
EOF

head -3 $Data_Dir/all.gct | tail -1| sed -re "s/ /\n/g" | tail -n +3| sed  's/\(.*\)/"\1",1/g' >> $Data_Dir/batches.csv
head $Data_Dir/batches.csv
```
### Correct format of all.csv
The file all.gct file needed modification. First column needs to have the sequence and gene name joined by an underscore.
```{r correct_csv_format}
table <- read.table(file.path(data_dir,"all.gct"),header=T, skip=2, check.names=F, as.is=T)
table[1:3,1:3]
table[,1] <- paste0(table[,1],'_',table[,2])
table[1:3,1:3]
outfile <- file.path(data_dir,"all.gct")
cat(header1, header2, file=outfile, sep="\n")
write.table(table,file=outfile, append=T, row.names=F, quote=F)
opts_knit$set(root.dir = opts_knit$get("demeter.dir"))
```

Preprocess the data from raw GCT files from Achilles into an R structure that DEMETER operates on
```{bash demeter_preprocess}
$Src_Dir/demeter/preprocess-gcts $Data_Dir/reformatted.Rdata $Data_Dir/batches.csv $Data_Dir/all.gct
```
Generate the final z-scored gene solutions, using the parameters used for the Achilles data 
```{bash demeter_final}
cd $Src_Dir/demeter
./eval-demeter-final --full.data.file=$Data_Dir/reformatted.Rdata \
     --dest.dir=$Data_Dir/final --learning.rate=0.005 --randseed=1 \
     --G.S=4e-5 --alpha.beta.gamma=0.9 --max.num.iter=250
